---
title: Practical 2
date: today
date-format: long
author: Melvil Deleage, Jeff Macaraeg
toc: true
format:
  pdf:
    code-fold: false
    toc: true
    code-block-border-left: "#0d6efd"
    code-block-bg: true
    highlight-style: github
    geometry: margin = 1.5cm
editor_options: 
  chunk_output_type: console
---

# Part 1 - Venice

```{r}
load("/Users/jeffmacaraeg/Documents/GitHub/Retake_RA/data/Practical2/niveau.Rdata")
```

## Question a) 

> Read in the data. Extract and represent the yearly max values from 1940 to 2009. What do you
observe ?

```{r}
library(tidyverse)
library(lubridate)
library(extRemes)

niveau$Date <- as.Date(niveau$Zeitstempel, format="%Y-%m-%d %H:%M:%S")

niveau <- niveau %>%
  mutate(Year = year(Date))

niveau_filtered <- niveau %>%
  filter(Year >= 1940, Year <= 2009)

venice_max <- niveau_filtered %>%
  group_by(Year) %>%
  summarise(Max_Sea_Level = max(Wert, na.rm = TRUE))

ggplot(venice_max, aes(x = Year, y = Max_Sea_Level)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Yearly max value (1940–2009)",
       x = "Year", y = "Yearly max (cm)")
```

There seems to be an increasing trend overtime. However, there is some variability from years to years, some of them indicating potential extreme events.

## Question b)

```{r}
# Fit a linear model on the observed data
lm_fit <- lm(Max_Sea_Level ~ Year, data = venice_max)
summary(lm_fit)  # Optional: review model details

# Create a new data frame for the prediction period
future_years <- data.frame(Year = 2010:2022)

# Obtain predictions with 95% confidence intervals
predictions <- predict(lm_fit, newdata = future_years, interval = "confidence")
pred_df <- cbind(future_years, as.data.frame(predictions))

# Plot the historical data, fitted model, and predictions with confidence intervals
ggplot() +
  # Historical data
  geom_line(data = venice_max, aes(x = Year, y = Max_Sea_Level), color = "blue") +
  geom_point(data = venice_max, aes(x = Year, y = Max_Sea_Level), color = "blue") +
  # Fitted linear model for the prediction period
  geom_line(data = pred_df, aes(x = Year, y = fit), color = "red", linetype = "dashed") +
  # Confidence intervals as a ribbon
  geom_ribbon(data = pred_df, aes(x = Year, ymin = lwr, ymax = upr), alpha = 0.2, fill = "red") +
  labs(title = "Yearly Maximum Sea Levels: Historical Data and Predictions (2010–2022)",
       x = "Year", y = "Max Sea Level (cm)") +
  theme_minimal()
```

The linear model shows the histrocial trend into the future. Depending on the fit, the prediction might show a contiuing increase/decrease in yearly maximum sea levels. Using a 95% confidence interval, we can see that the interval tend to widen, which might reflect some uncertainties in the predictions.

## Question c)

```{r}
# These are the individual observations (from the full dataset) with sea levels > 140 cm.
observed_high <- niveau %>% filter(Wert > 140)

ggplot() +
  # Historical yearly maximum sea levels (1940–2009)
  geom_line(data = venice_max, aes(x = Year, y = Max_Sea_Level), color = "blue") +
  geom_point(data = venice_max, aes(x = Year, y = Max_Sea_Level), color = "blue") +
  
  # Predicted values for 2010–2022 with a dashed line.
  geom_line(data = pred_df, aes(x = Year, y = fit), color = "red", linetype = "dashed") +
  
  # Confidence interval for predictions as a shaded ribbon.
  geom_ribbon(data = pred_df, aes(x = Year, ymin = lwr, ymax = upr), fill = "red", alpha = 0.2) +
  
  # Observed sea level values greater than 140 cm, plotted as distinct points.
  geom_point(data = observed_high, aes(x = Year, y = Wert), color = "darkgreen", shape = 4, size = 3) +
  
  labs(title = "Yearly Maximum Sea Levels and Observations > 140 cm",
       subtitle = "Historical (1940–2009), Predicted (2010–2022), and Individual Observations > 140 cm",
       x = "Year", y = "Sea Level (cm)") +
  theme_minimal()
```

## Question d)

```{r}
# (a) GEV with constant parameters
gev_const <- fevd(venice_max$Max_Sea_Level, data = venice_max, type = "GEV", method = "MLE")
summary(gev_const)

# (b) GEV with time-varying location (linear in Year)
gev_tv <- fevd(venice_max$Max_Sea_Level, data = venice_max, type = "GEV", method = "MLE",
               location.fun = ~ Year)
summary(gev_tv)

# 3. Likelihood Ratio Test
LRT_stat <- 2 * (logLik(gev_tv) - logLik(gev_const))
p_val <- 1 - pchisq(LRT_stat, df = 1)
cat("LRT statistic:", LRT_stat, "\n", "p-value:", p_val, "\n")

# 4. Diagnostic Plots
par(mfrow = c(2, 2))
plot(gev_const)
title("Constant-Param GEV", outer = TRUE)

par(mfrow = c(2, 2))
plot(gev_tv)
title("Time-Varying Location GEV", outer = TRUE)
par(mfrow = c(1, 1))  # reset

# 5. Compute 13-year Return Levels for 2010–2022
future_data <- data.frame(Year = 2010:2022)
coef_tv <- coef(gev_tv)
# For GEV: mu(Year) = β0 + β1*Year, scale = σ, shape = ξ
future_data <- future_data %>%
  mutate(
    mu = coef_tv["locationIntercept"] + coef_tv["locationYear"] * Year,
    sigma = coef_tv["scale"],
    xi = coef_tv["shape"],
    R13 = mu - (sigma / xi) * (1 - (-log(1 - 1/13))^(-xi))
  )

print(future_data)
```




# Part 2 - Nuclear Reactors

## Question a)

```{r}

```

