---
title: Practical 1
date: today
date-format: long
author: Melvil Deleage, Jeff Macaraeg
toc: true
format:
  pdf:
    code-fold: false
    toc: true
    code-block-border-left: "#0d6efd"
    code-block-bg: true
    highlight-style: github
    geometry: margin = 1.5cm
editor_options: 
  chunk_output_type: console
---

# Part 1 - Financial Returns & Normality

## Question a)

> Read in the data. Then, assess the stationarity of the (raw) stock indices.

SP500

```{r}
library(QRM)
library(tseries)

# Load the SP500 data
data(sp500)

# Take a quick look at the SP500 data
head(sp500)

# Plot the SP500 index to visually assess stationarity
plot(sp500, type = "l", col = "blue", main = "SP500 Index", ylab = "Index Value")

# Augmented Dickey-Fuller (ADF) test to check stationarity
adf_test_sp500 <- adf.test(sp500)

# Print the results of the ADF test
print(adf_test_sp500)
```

As a result, we can see that the p-value = 0.9044, which is higher than 0.05. This means that the series is not stationary.

CAC50

```{r}
# Load the NIKKEI data
data("cac40")

# Plot the NIKKEI index to visually assess stationarity
plot(cac40, type = "l", col = "blue", main = "CAC40 Index", ylab = "Index Value")

# Augmented Dickey-Fuller (ADF) test to check stationarity
adf_test_cac40 <- adf.test(cac40)

# Print the results of the ADF test
print(adf_test_cac40)
```

Just like for SP500, the p-value is 0.9676 which is higher than 0.05, so the series is not stationary.

## Question b)

> Create a function to transform the daily stock indices into their daily negative log returns
counterparts. Plot the latter series and assess their stationarity. To compare the series, also plot the negative log returns on a common scale to all indices.

```{r}
library(ggplot2)

# Define a function to compute negative log returns, handling NA values
negative_log_returns <- function(index_data) {
  index_data <- na.omit(index_data)  # Remove NAs
  return(-diff(log(index_data)))
}

# Check structure of the data
print(str(cac40))
print(str(sp500))

# Compute negative log returns for CAC 40 and SP500
neg_log_returns_cac40 <- negative_log_returns(as.numeric(cac40))  # Ensure numeric
neg_log_returns_sp500 <- negative_log_returns(as.numeric(sp500))  # Ensure numeric

# Plot individual negative log return series
par(mfrow = c(2, 1))  # Arrange plots in 2 rows, 1 column

plot(neg_log_returns_cac40, type = "l", col = "red", main = "Negative Log Returns - CAC 40", ylab = "Log Returns")
plot(neg_log_returns_sp500, type = "l", col = "blue", main = "Negative Log Returns - SP500", ylab = "Log Returns")

# Reset plot layout
par(mfrow = c(1, 1))

# Plot both indices on the same scale
plot(neg_log_returns_cac40, type = "l", col = "red", 
     ylim = range(c(neg_log_returns_cac40, neg_log_returns_sp500), na.rm = TRUE),
     main = "Comparison of Negative Log Returns", ylab = "Log Returns", xlab = "Time")
lines(neg_log_returns_sp500, col = "blue")
legend("topright", legend = c("CAC40", "SP500"), col = c("red", "blue"), lty = 1)

# Perform ADF test for stationarity
adf_cac40 <- adf.test(neg_log_returns_cac40)
adf_sp500 <- adf.test(neg_log_returns_sp500)

# Print ADF test results
print("ADF Test for CAC 40:")
print(adf_cac40)
print("ADF Test for SP500:")
print(adf_sp500)
```

As both p-values (p-value = 0.01) are lower than 0.05, we reject the hypothesis and thus, both series are stationary.

## Question c)

> Draw histograms of the negative log returns and compare them to the Normal distribution.
What do you observe?

```{r}
# Set up plotting area (2 rows, 1 column)
par(mfrow = c(2, 1))

# Histogram for CAC 40
hist(neg_log_returns_cac40, breaks = 50, probability = TRUE, col = "red", border = "black",
     main = "Histogram of Negative Log Returns - CAC 40", xlab = "Log Returns", ylab = "Density")
curve(dnorm(x, mean = mean(neg_log_returns_cac40, na.rm = TRUE), 
            sd = sd(neg_log_returns_cac40, na.rm = TRUE)), 
      col = "blue", lwd = 2, add = TRUE)  # Overlay Normal distribution

# Histogram for SP500
hist(neg_log_returns_sp500, breaks = 50, probability = TRUE, col = "blue", border = "black",
     main = "Histogram of Negative Log Returns - SP500", xlab = "Log Returns", ylab = "Density")
curve(dnorm(x, mean = mean(neg_log_returns_sp500, na.rm = TRUE), 
            sd = sd(neg_log_returns_sp500, na.rm = TRUE)), 
      col = "red", lwd = 2, add = TRUE)  # Overlay Normal distribution

# Reset plot layout
par(mfrow = c(1, 1))
```

Both deviate from the Normal distribution, with higher peaks and fatter tails which suggests more extreme values than we expect under normality.

## Question d)

> Check the normality assumption of the negative log returns using QQ-plots. What is your
conclusion?

```{r}
# Set up plotting area
par(mfrow = c(2, 1))

# QQ-Plot for CAC 40
qqnorm(neg_log_returns_cac40, main = "QQ-Plot of Negative Log Returns - CAC 40", col = "red")
qqline(neg_log_returns_cac40, col = "blue", lwd = 2)

# QQ-Plot for SP500
qqnorm(neg_log_returns_sp500, main = "QQ-Plot of Negative Log Returns - SP500", col = "blue")
qqline(neg_log_returns_sp500, col = "red", lwd = 2)

# Reset plot layout
par(mfrow = c(1, 1))
```

Neg log returns are not normally distributed, as the points deviate from the line.

## Question e)

> Formally test the normality assumption of the negative log returns using an Anderson-Darling
testing procedure. Do you reject the Normal hypothesis?

```{r}
# Load necessary library
library(nortest)  

# Perform Anderson-Darling test for CAC 40
ad_test_cac40 <- ad.test(neg_log_returns_cac40)

# Perform Anderson-Darling test for SP500
ad_test_sp500 <- ad.test(neg_log_returns_sp500)

# Print test results
cat("📊 Anderson-Darling Test for CAC 40:\n")
print(ad_test_cac40)

cat("\n📊 Anderson-Darling Test for SP500:\n")
print(ad_test_sp500)
```

As both p-values are lower than 0.05, we reject the null hypothesis, meaning that the returns are not normally distributed.

## Question f)

> Use the fitdistr() function from the MASS package in order to obtain the (maximum-likelihood
estimated) parameters of distributions you could imagine for the negative log returns. Try to fit
at least two diﬀerent distributions on the data and, using an information criteria (such as the
AIC), decide which distributional framework fits best for each of the series.

```{r}
# Load necessary libraries
library(MASS)

# Ensure there are no non-positive values before fitting a log-normal distribution
# Take the absolute value of the negative log returns for log-normal fitting
# Note: Use this only for log-normal fitting, not for other distributions
neg_log_returns_cac40_abs <- abs(neg_log_returns_cac40)
neg_log_returns_sp500_abs <- abs(neg_log_returns_sp500)

# Fit Normal distribution to CAC 40 and SP500 log returns
fit_normal_cac40 <- fitdistr(neg_log_returns_cac40, densfun = "normal")
fit_normal_sp500 <- fitdistr(neg_log_returns_sp500, densfun = "normal")

# Fit Log-normal distribution to CAC 40 and SP500 absolute log returns
fit_lognormal_cac40 <- fitdistr(neg_log_returns_cac40_abs, densfun = "lognormal")
fit_lognormal_sp500 <- fitdistr(neg_log_returns_sp500_abs, densfun = "lognormal")

# Fit Gamma distribution to CAC 40 and SP500 log returns
fit_gamma_cac40 <- fitdistr(neg_log_returns_cac40, densfun = "gamma")
fit_gamma_sp500 <- fitdistr(neg_log_returns_sp500, densfun = "gamma")

# Optionally, fit t-distribution (for heavy tails)
fit_t_cac40 <- fitdistr(neg_log_returns_cac40, densfun = "t")
fit_t_sp500 <- fitdistr(neg_log_returns_sp500, densfun = "t")

# Print the fit results and AIC values for comparison
cat("Fit Results for CAC 40:\n")
cat("Normal AIC:", fit_normal_cac40$aic, "\n")
cat("Lognormal AIC:", fit_lognormal_cac40$aic, "\n")
cat("Gamma AIC:", fit_gamma_cac40$aic, "\n")
cat("T-distribution AIC:", fit_t_cac40$aic, "\n")

cat("\nFit Results for SP500:\n")
cat("Normal AIC:", fit_normal_sp500$aic, "\n")
cat("Lognormal AIC:", fit_lognormal_sp500$aic, "\n")
cat("Gamma AIC:", fit_gamma_sp500$aic, "\n")
cat("T-distribution AIC:", fit_t_sp500$aic, "\n")
```

ne marche pas

## Question g)

> If this has not been done in (f), fit a t-distribution to the negative log returns using fitdistr().
Using a QQ-plot for each of the series, decide whether the fit is better than with a Normal
distribution, based on your answer in (d).

xxx

# Part 2 - Financial time series, volatility and the random walk hypothesis

## Question a)

> Plot the ACF of all the series in Part 1 (i.e. the raw series as well as the negative log returns).
What do you observe?

```{r}
# Compute negative log returns
negative_log_returns <- function(index_data) {
  index_data <- na.omit(index_data)  # Remove missing values
  return(-diff(log(index_data)))
}

neg_log_returns_cac40 <- negative_log_returns(as.numeric(cac40))
neg_log_returns_sp500 <- negative_log_returns(as.numeric(sp500))

# Set up plotting area (2 rows, 2 columns)
par(mfrow = c(2, 2))

# ACF for raw CAC40 series
acf(cac40, main = "ACF of CAC40 Raw Series", col = "red")

# ACF for raw SP500 series
acf(sp500, main = "ACF of SP500 Raw Series", col = "blue")

# ACF for negative log returns of CAC40
acf(neg_log_returns_cac40, main = "ACF of CAC40 Negative Log Returns", col = "red")

# ACF for negative log returns of SP500
acf(neg_log_returns_sp500, main = "ACF of SP500 Negative Log Returns", col = "blue")

# Reset plot layout
par(mfrow = c(1, 1))
```

The raw series show high autocorrelation, so the stock prices follow trends. For the negative log returns, it shows little to no autocorrelation, suggesting that the daily returns are close to uncorrelated.

## Question b)

> Use a Ljung-Box procedure to formally test for (temporal) serial dependence in the series. What
is your conclusion?

```{r}
# Load necessary library
library(stats)  

# Perform Ljung-Box test for raw CAC40 series
ljung_raw_cac40 <- Box.test(cac40, lag = 20, type = "Ljung-Box")

# Perform Ljung-Box test for raw SP500 series
ljung_raw_sp500 <- Box.test(sp500, lag = 20, type = "Ljung-Box")

# Perform Ljung-Box test for CAC40 negative log returns
ljung_nlr_cac40 <- Box.test(neg_log_returns_cac40, lag = 20, type = "Ljung-Box")

# Perform Ljung-Box test for SP500 negative log returns
ljung_nlr_sp500 <- Box.test(neg_log_returns_sp500, lag = 20, type = "Ljung-Box")

# Print results
cat("📊 Ljung-Box Test for CAC40 Raw Series:\n")
print(ljung_raw_cac40)

cat("\n📊 Ljung-Box Test for SP500 Raw Series:\n")
print(ljung_raw_sp500)

cat("\n📊 Ljung-Box Test for CAC40 Negative Log Returns:\n")
print(ljung_nlr_cac40)

cat("\n📊 Ljung-Box Test for SP500 Negative Log Returns:\n")
print(ljung_nlr_sp500)
```

For the raw series for both CAC40 and SP500, we obtain a p-value that is lower than 0.05, so we reject the null hypothesis and the raw series show autocorrelation. For both negative log returns, the p-values are higher than the raw series, but still smaller than the p-value, so again they show autocorrelation.

## Question c)

> Propose ARIMA models for each of the negative log returns series, based on visualisation tools
(e.g. ACF and PACF). Select an ARIMA model using auto.arima() (forecast package) to
each of the negative log returns series. Comment on the diﬀerence. Assess the residuals of the
resulting models.

```{r}
# Load necessary libraries
library(forecast)  

# Fit ARIMA models using auto.arima()
arima_cac40 <- auto.arima(neg_log_returns_cac40)
arima_sp500 <- auto.arima(neg_log_returns_sp500)

# Print selected models
cat("📊 Selected ARIMA Model for CAC40:\n")
print(arima_cac40)

cat("\n📊 Selected ARIMA Model for SP500:\n")
print(arima_sp500)

# Plot ACF & PACF to visualize possible ARIMA orders
par(mfrow = c(2, 2))

acf(neg_log_returns_cac40, main = "ACF - CAC40 Negative Log Returns", col = "red")
pacf(neg_log_returns_cac40, main = "PACF - CAC40 Negative Log Returns", col = "red")

acf(neg_log_returns_sp500, main = "ACF - SP500 Negative Log Returns", col = "blue")
pacf(neg_log_returns_sp500, main = "PACF - SP500 Negative Log Returns", col = "blue")

par(mfrow = c(1, 1))

# Assess residuals of the ARIMA models
par(mfrow = c(2, 2))

checkresiduals(arima_cac40)
checkresiduals(arima_sp500)

par(mfrow = c(1, 1))
```

conclusion

## Question d)

> Assess the residuals of the resulting models from (c), both their raw values and their absolute
values, through visual tools (such as the ACF) and formal tests (e.g. Ljung-Box). What do you
conclude about the independence assumption?

```{r}
# Extract residuals from ARIMA models
residuals_cac40 <- residuals(arima_cac40)
residuals_sp500 <- residuals(arima_sp500)

# Compute absolute residuals
abs_residuals_cac40 <- abs(residuals_cac40)
abs_residuals_sp500 <- abs(residuals_sp500)

# Set up plotting area (2 rows, 2 columns)
par(mfrow = c(2, 2))

# ACF of raw residuals
acf(residuals_cac40, main = "ACF - CAC40 Residuals", col = "red")
acf(residuals_sp500, main = "ACF - SP500 Residuals", col = "blue")

# ACF of absolute residuals (to check volatility clustering)
acf(abs_residuals_cac40, main = "ACF - CAC40 Absolute Residuals", col = "red")
acf(abs_residuals_sp500, main = "ACF - SP500 Absolute Residuals", col = "blue")

par(mfrow = c(1, 1))

# Perform Ljung-Box tests (raw residuals)
ljung_residuals_cac40 <- Box.test(residuals_cac40, lag = 20, type = "Ljung-Box")
ljung_residuals_sp500 <- Box.test(residuals_sp500, lag = 20, type = "Ljung-Box")

# Perform Ljung-Box tests (absolute residuals)
ljung_abs_residuals_cac40 <- Box.test(abs_residuals_cac40, lag = 20, type = "Ljung-Box")
ljung_abs_residuals_sp500 <- Box.test(abs_residuals_sp500, lag = 20, type = "Ljung-Box")

# Print Ljung-Box test results
cat("📊 Ljung-Box Test for CAC40 Raw Residuals:\n")
print(ljung_residuals_cac40)

cat("\n📊 Ljung-Box Test for SP500 Raw Residuals:\n")
print(ljung_residuals_sp500)

cat("\n📊 Ljung-Box Test for CAC40 Absolute Residuals:\n")
print(ljung_abs_residuals_cac40)

cat("\n📊 Ljung-Box Test for SP500 Absolute Residuals:\n")
print(ljung_abs_residuals_sp500)
```

## Question e)

> Plot the volatility of the raw series of indices. What is your conclusion on the homoscedasticity
assumption?

```{r}
# Load required library
library(TTR)  # For volatility estimation

# Compute rolling standard deviation as a measure of volatility
volatility_cac40 <- runSD(cac40, n = 30)  # 30-day rolling window
volatility_sp500 <- runSD(sp500, n = 30)  # 30-day rolling window

# Plot the volatility of CAC40 and SP500
par(mfrow = c(2, 1))

plot(volatility_cac40, type = "l", col = "red", 
     main = "Volatility of CAC40 (30-day Rolling SD)", ylab = "Volatility")
plot(volatility_sp500, type = "l", col = "blue", 
     main = "Volatility of SP500 (30-day Rolling SD)", ylab = "Volatility")

par(mfrow = c(1, 1))
```

## Question f)

> Residual serial correlation can be present when fitting a GARCH directly on the negative log
returns. Hence, in order to circumvent this problem, it is possible to use the following two-step
approach:
• fit an ARMA(p,q) on the negative log returns;
• fit a GARCH(1,1) on the residuals of the ARMA(p,q) fit.
Proceed with the above recipe. Assess the quality of the above fit.

```{r}
# Load required libraries
library(forecast)  # For ARMA model
library(rugarch)   # For GARCH model

# Step 1: Fit ARMA model on negative log returns
arma_cac40 <- auto.arima(neg_log_returns_cac40, max.p = 5, max.q = 5, seasonal = FALSE, ic = "AIC")
arma_sp500 <- auto.arima(neg_log_returns_sp500, max.p = 5, max.q = 5, seasonal = FALSE, ic = "AIC")

# Extract residuals
residuals_cac40 <- residuals(arma_cac40)
residuals_sp500 <- residuals(arma_sp500)

# Step 2: Fit GARCH(1,1) model on ARMA residuals
garch_spec <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
                         mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
                         distribution.model = "norm")

garch_cac40 <- ugarchfit(spec = garch_spec, data = residuals_cac40)
garch_sp500 <- ugarchfit(spec = garch_spec, data = residuals_sp500)

# Print GARCH model summary
cat("📊 GARCH(1,1) Model for CAC40:\n")
print(garch_cac40)

cat("\n📊 GARCH(1,1) Model for SP500:\n")
print(garch_sp500)

# Plot standardized residuals
par(mfrow = c(2, 2))

plot(garch_cac40@fit$residuals, type = "l", col = "red", main = "Standardized Residuals - CAC40")
plot(garch_sp500@fit$residuals, type = "l", col = "blue", main = "Standardized Residuals - SP500")

# ACF of standardized residuals
acf(garch_cac40@fit$residuals, main = "ACF of GARCH Residuals - CAC40", col = "red")
acf(garch_sp500@fit$residuals, main = "ACF of GARCH Residuals - SP500", col = "blue")

par(mfrow = c(1, 1))

# Perform Ljung-Box test to check for serial correlation
ljung_cac40 <- Box.test(garch_cac40@fit$residuals, lag = 20, type = "Ljung-Box")
ljung_sp500 <- Box.test(garch_sp500@fit$residuals, lag = 20, type = "Ljung-Box")

cat("\n📊 Ljung-Box Test for GARCH(1,1) Residuals - CAC40:\n")
print(ljung_cac40)

cat("\n📊 Ljung-Box Test for GARCH(1,1) Residuals - SP500:\n")
print(ljung_sp500)
```

